---
title: "Bayesian Underreported Count Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{css, include = FALSE}
.justify {
  text-align: justify;
}
```

The **BUCM** packages provides an interface to fit Bayesian model to count data. The package is in it's simplest form without allowing for covariates(coming soon).

Three different distributions of count data are supported. However, **BUCM** also includes model selection components that users can use to determine plausible underlying distribution for the count data. Model fit can easily be assess with other diagnostic functions available in the package. To the best of our knowledge, other MCMC diagnostic tools support this package.

## Installation

Install the latest version from **Github**

```{r, eval = FALSE}
remotes::install_github("TheoAnim/BUCM")
```

## How to use BUCM
```{r setup}
library(BUCM)
library(ggplot2)
```


As an example, we simulate data from Poisson subject to underreporting. Assume some validation data through which we know the true(latent) event count denoted by
$$y^{\ast} \sim Poisson(\lambda)$$ 
where $\lambda > 0$ is the expected(true) count.

For all units, the reported count $y$ arises from a binomial mechanism. Conditional on the true count $y^{\ast}$,
$$y|y^{\ast} \sim Binomial(y^{\ast}, p)$$ 
where $p$ denotes the reporting probability.

For the remaining units(the main study sample), only the underreported counts $y$ are observed.
Marginally, we can derived the distribution of $y$ as:
$$y \sim Poisson(\lambda p)$$ 

```{r}
set.seed(3)
nv <- 30
nobs <- 100
lambda <- 4
p <- .6
ystar <- rpois(nv, lambda)
yval <- rbinom(nv, ystar, p)
yobs <- rpois(nobs, lambda * p)


data.frame(y = c(ystar, yval, yobs),
           type = c(rep("ystar", nv), rep("yval", nv), rep("yobs", nobs))) |>
  ggplot() +
  geom_bar(aes(y)) +
  facet_grid( ~ type)+
  labs(x = "data")
```


```{r}
jags_models <- BUCM::urc_mcmc(x = list(ystar = ystar, yval = yval, yobs = yobs))
jags_models$dic_best

```

The log-likelihoods are monitored for WAIC and LOO-CV computation, so you must explicitly specify the parameters; otherwise, a summary of the log-likelihoods will be included. `ucr_mcmc` also returns the plausible model if using WAIC as the model selection method. The likelihoods are also returned for each model allowing users to compute and diagnose WAIC and LOO-CV methods(Take note of warnings for waic computations).
```{r}
jags_models$models$poisson |> MCMCvis::MCMCsummary(params = c("lambda", "p"))
```

```{r}
jags_models$models$poisson |> BUCM::urc_trace(parameters = c("mu", "lambda", "p"))
```

```{r}
jags_models$models$poisson$BUGSoutput$sims.array |> bayesplot::mcmc_dens(pars = c("mu", "lambda", "p"))
```


```{r}
jags_models$models$poisson$BUGSoutput$sims.array |> bayesplot::mcmc_pairs(pars = c("mu", "lambda", "p"))
```


Assuming no validation set, we can try informative priors in the parameters

```{r}
jags_model2 <- BUCM::urc_mcmc(
  x = list(ystar = NA, yval = NA, yobs = yobs),
  prior_lambda = "dgamma(40, 10)",
  prior_p = "dbeta(60, 40)"
)


jags_model2$dic_best


jags_model2$models$poisson |> MCMCvis::MCMCsummary(params = c("lambda", "p"))
jags_model2$models$poisson$BUGSoutput$sims.array |> bayesplot::mcmc_trace(pars = c("lambda", "p"))

jags_model2$models$poisson$BUGSoutput$sims.array |> bayesplot::mcmc_dens(pars = c("lambda", "p"))
```


