---
title: "Bayesian Underreported Count Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Underreported Count Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
)
bayesplot::color_scheme_set("viridis")
```

```{css, include = FALSE}
.justify {
  text-align: justify;
}
```

The **BUCM** packages provides an interface to fit Bayesian model to count data. The package is in it's simplest form without allowing for covariates(coming soon).

Three different distributions of count data are supported. However, **BUCM** also includes model selection components that users can use to determine plausible underlying distribution for the count data. To the best of our knowledge, other MCMC diagnostic tools such as bayesplot.

## Installation

Install the latest version from **Github**

```{r, eval = FALSE}
remotes::install_github("TheoAnim/BUCM")
```

## Libraries
```{r setup, warning = FALSE, message = FALSE}
library(BUCM)
library(tidyverse)
library(loo)
```

## Illustrative Example

As an example, we simulate data from Poisson subject to underreporting. Assume some validation data through which we know the true(latent) event count denoted by
$$Y^{\ast} \sim Poisson(\lambda)$$ 
where $\lambda > 0$ is the expected(true) count.

For all units, the reported count $Y$ arises from a binomial mechanism. Conditional on the true count $T^{\ast}$,
$$Y|y^{\ast} \sim Binomial(y^{\ast}, p)$$ 
where $p$ denotes the reporting probability.

For the remaining units(the main study sample), only the underreported counts $Y$ are observed.
Marginally, we can derived the distribution of $y$ as:
$$Y \sim Poisson(\lambda p)$$ 

```{r}
set.seed(3)
nv <- 30
nobs <- 100
lambda <- 4
p <- .6
ystar <- rpois(nv, lambda)
yval <- rbinom(nv, ystar, p)
yobs <- rpois(nobs, lambda * p)


data.frame(y = c(ystar, yval, yobs),
           type = c(rep("ystar", nv), rep("yval", nv), rep("yobs", nobs))
) |>
  ggplot() +
  geom_bar(aes(y)) +
  facet_grid( ~ type)+
  labs(x = "data")
```


## Fit Model

With validation data available, prior specifications can be noninformative, weakly informative or informative.

```{r}
bayes_count_models <- BUCM::urc_mcmc(x = list(ystar = ystar, yval = yval, yobs = yobs), 
                                    prior_p = "dbeta(3, 2)", 
                                    prior_lambda = "dgamma(4, 1)")
```


### Model Diagnostics

The log-likelihoods are monitored for WAIC and PSIS-LOO computations, so you must explicitly specify the parameters; otherwise, a summary of the log-likelihoods will be included. `ucr_mcmc` also returns the plausible model if using DIC/WAIC/PSIS-LOO as the model selection method. The likelihoods are also returned for each model allowing users to compute and diagnose WAIC and PSIS-LOO methods (Take note of warnings for WAIC and PSIS-LOO computations, see original `loo` documentation).


```{r}
bayes_count_models$models$poisson |> 
  BUCM::urc_trace(parameters = c("mu", "lambda", "p"))
```

**BUCM** is compatible with bayesplot package. Pass array of posterior draws to create trace plots, density, posterior intervals and other diagnostic visualizations.

```{r}
jm_poisson_array <- bayes_count_models$models$poisson$BUGSoutput$sims.array
bayesplot::mcmc_dens(jm_poisson_array, pars = c("mu", "lambda", "p"))
```


```{r}
bayesplot::mcmc_pairs(jm_poisson_array, pars = c("mu", "lambda", "p"))
```

### Posterior Summaries
```{r}
bayes_count_models$models$poisson |> 
  MCMCvis::MCMCsummary(params = c("lambda", "p"))
```

### Model Selection

```{r}
bayes_count_models$models$poisson |> 
  MCMCvis::MCMCsummary(params = c("lambda", "p"))
```


## No validation data

Assuming no validation set, we can try informative priors on the parameters

```{r, eval = FALSE}
bayes_count_models2 <- BUCM::urc_mcmc(
  x = list(ystar = NA, yval = NA, yobs = yobs),
  prior_lambda = "dgamma(16, 4)",
  prior_p = "dbeta(60, 40)"
)
```

