---
title: "Bayesian Underreported Count Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Underreported Count Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%"
)
```

```{r}
library(ggplot2)
```

```{css, include = FALSE}
.justify {
  text-align: justify;
}
```

**BUCM** (Bayesian Underreported Count Models) is an R package for Bayesian
analysis of count data with and without underreporting. The package provides
one-sample and two-sample Poisson, zero-inflated Poisson (ZIP), and negative
binomial models.

BUCM includes both naive models, which assume no underreporting, and
underreported models that account for incomplete reporting. Model estimation
is performed using Markov chain Monte Carlo (MCMC) methods through JAGS.

Model comparison is available using DIC, WAIC, and PSIS-LOO.

## Installation

Install the latest version from **Github**

```{r, eval = FALSE}
install.packages("pak")
pak::pkg_install("TheoAnim/BUCM")
```

## Illustrative Example

As an example, we simulate data from Poisson subject to underreporting. Assume some validation data through which we know the true(latent) event count denoted by
$$Y^{\ast} \sim Poisson(\lambda)$$ 
where $\lambda > 0$ is the expected(true) count.

For all units, the reported count $Y$ arises from a binomial mechanism. Conditional on the true count $T^{\ast}$,
$$Y|y^{\ast} \sim Binomial(y^{\ast}, p)$$ 
where $p$ denotes the reporting probability.

For the remaining units(the main study sample), only the underreported counts $Y$ are observed.
Marginally, we can derived the distribution of $y$ as:
$$Y \sim Poisson(\lambda p)$$ 

## Generate the Data

```{r}
set.seed(123)

# Simulation settings
lambda <- 5
p <- 0.7
nv <- 100      # Validation sample size
nobs <- 300    # Non-validation sample size

# Validation sample: latent counts (Y*) and observed counts (Y)
ystar <- rpois(
  nv,
  lambda = lambda
)

yval <- rbinom(
  nv,
  size = ystar,
  prob = p
)

# Observed counts subject to underreporting
yobs <- rpois(
  nobs,
  lambda = lambda * p
)
```

Here, `ystar` contains the true counts in the validation sample, `yval`
contains the corresponding reported counts generated through the binomial
reporting mechanism, and `yobs` contains the observed counts subject to
underreporting in the main study sample.

```{r}
data.frame(
  y = c(ystar, yval, yobs),
  type = c(
    rep("ystar", nv),
    rep("yval", nv),
    rep("yobs", nobs)
  )
) |>
  ggplot() +
  geom_bar(aes(y)) +
  facet_grid(~ type) +
  labs(
    x = "Count",
    y = "Frequency"
  )
```


## Fit Model

With validation data available, prior specifications can be excluded, noninformative, weakly informative or informative.

```{r}
set.seed(123)
bayes_count_models <- BUCM::urc_mcmc(
  x = list(ystar = ystar, yval = yval, yobs = yobs),
  prior_p = "dbeta(22, 18)",
  prior_lambda = "dgamma(4, 1)",
  thresh = 5
)
```

```{r}
ls(bayes_count_models)
```

### Model Diagnostics

BUCM provides extra diagnostics functions `urc_trace`, `urc_acf`, `urc_density` and `urc_hist`.
`urc_mcmc` also returns the plausible model if using DIC/WAIC/PSIS-LOO as the model selection method. The likelihoods are also returned for each model allowing users to compute and diagnose WAIC and PSIS-LOO methods (Take note of warnings for WAIC and PSIS-LOO computations, see original `loo` documentation).


```{r}
bayes_count_models$models$poisson |> 
  BUCM::urc_trace(parameters = c("lambda", "p"), alpha = .5)
```

**BUCM** is compatible with bayesplot package. Pass array of posterior draws to create trace plots, density, posterior intervals and other diagnostic visualizations. The log-likelihoods are monitored for WAIC and PSIS-LOO computations, so you must explicitly specify the parameters; otherwise, log-likelihoods will be included. 
```{r}
jm_poisson_array <- bayes_count_models$models$poisson$BUGSoutput$sims.array
bayesplot::mcmc_dens(jm_poisson_array, pars = c("mu", "lambda", "p"))
```

```{r}
bayesplot::mcmc_pairs(jm_poisson_array, pars = c("mu", "lambda", "p"))
```

### Posterior Summaries

```{r}
bayes_count_models$models$poisson |>
  BUCM::urc_posterior_summary(exclude = c("mu", "deviance"), digits = 2)
```

Posterior summaries can also be obtained using `MCMCvis`.
```{r}
bayes_count_models$models$poisson |> 
  MCMCvis::MCMCsummary(params = c("lambda", "p"))
```

### Model Selection

(a) Return values for each criterion for each model
```{r}
# using DIC for model selection
bayes_count_models$dics |> 
  kableExtra::kable(caption = "DIC Comparison")
```

```{r}
# using WAIC for model selection
bayes_count_models$waics |> 
  kableExtra::kable(caption = "WAIC Comparison")
```

```{r}
# using PSIS-LOO for model selection
bayes_count_models$loos |> 
  kableExtra::kable(caption = "PSIS-LOO Comparison")
```


(b) Show plausible model selected as being parsimonious or having the smallest criterion value

```{r}
data.frame(
  Model = c("DIC", "WAIC", "LOO"),
  "Best Selected" = c(
    bayes_count_models$dic_best,
    bayes_count_models$waic_best,
    bayes_count_models$loo_best
  )
) |>
  kableExtra::kable(caption = "Plausible model by each selection criterion")
```


## No validation data

Assuming no validation set, we can try informative priors on the parameters

```{r, eval = FALSE}
bayes_count_models2 <- BUCM::urc_mcmc(
  x = list(ystar = NA, yval = NA, yobs = yobs),
  prior_p = "dbeta(70, 30)",
  prior_lambda = "dgamma(25, 5)"
)
```

# Two-Sample Models

For the two-sample setting, we consider two independent Poisson populations with different underlying mean counts and reporting probabilities. No validation data are assumed to be available, so the model is fitted to the two observed underreported samples using informative priors centered on the values used to generate the data.
```{r}
set.seed(123)

# Simulation settings
lambda1 <- 5
lambda2 <- 8

p1 <- 0.7
p2 <- 0.6

nobs <- 300 # Non-validation sample size

# Observed counts subject to underreporting
yobs1 <- rpois(
  nobs,
  lambda = lambda1 * p1
)

yobs2 <- rpois(
  nobs,
  lambda = lambda2 * p2
)

```


## Fit the two-sample model without validation data
```{r}
bayes_two_count_models <- BUCM::urc_two_mcmc(
  x = list(
    yobs1 = yobs1,
    yobs2 = yobs2,
    ystar1 = NA,
    yval1 = NA,
    ystar2 = NA,
    yval2 = NA
  ),
  prior_p1 = "dbeta(70, 30)",
  prior_p2 = "dbeta(60, 40)",
  prior_lambda1 = "dgamma(25, 5)",
  prior_lambda2 = "dgamma(32, 4)"
)
```
```{r}
ls(bayes_two_count_models)
```


```{r}
bayes_two_count_models$models$poisson |> BUCM::urc_trace()
```

See the `urc_mcmc_naive()` and `urc_two_mcmc_naive` documentation for the required of the naive case which assumes no underreporting.

# Citation

If you use BUCM in published research, please cite:

> Anim Bediako, T., Roberman, J. L., Barth, J., & Stamey, J. D. (2026).
> *BUCM: An R package for Bayesian modeling of one and two sample count
> models with underreporting.*


